CF Registry Analysis

### 2025-08-05 Christine Chan/CF Registry/cf_registry_survival_analysis.qmd
HbA1c table for Monique

```{r}
#| include: false
library(Hmisc)
library(zoo)
library(tidyverse)
library(childsds)
library(gt)
library(gtsummary)
library(forcats)
library(labelled)
library(survival)
library(survminer)
library(knitr)
library(eulerr)
library(broom)
library(emmeans)
library(ggstats)
library(patchwork)
home_dir <- switch(Sys.info()["sysname"],
  "Darwin" = "/Users/timvigers/Library/CloudStorage/OneDrive-TheUniversityofColoradoDenver/Vigers/CF/Christine Chan/CF Registry",
  "Linux" = "/home/timvigers/OneDrive/Vigers/CF/Christine Chan/CF Registry"
)
github_dir <- switch(Sys.info()["sysname"],
  "Darwin" = "/Users/timvigers/Documents/GitHub",
  "Linux" = "/home/timvigers/Documents/GitHub"
)
knitr::opts_knit$set(root.dir = home_dir)
```

@@

  "A_pulmonarycomplications1", "A_mycocultureresults3"

@@

```{r}
# Per reviewer, out of X individuals with a diagnosis of CFRD, X number had an
# OGTT (and X had an A1c, but no OGTT) that year or the year prior.
# This is super slow but idc idc
t <- apply(first_cfrd, 1, function(r) {
  id <- as.numeric(r["eDWID"])
  year <- as.numeric(r["first_cfrd"])
  d <- annual[annual$eDWID == id & annual$ReviewYear %in% c(year, year - 1), ]
  if (any(!is.na(d$A_hgba1c))) {
    a1c <- T
  } else {
    a1c <- F
  }
  if (any(!is.na(d["A_twohour_bloodglucose"]) |
    !is.na(d["A_ogttfast_bloodglucose"]))) {
    ogtt <- T
  } else {
    ogtt <- F
  }
  if (a1c & ogtt) {
    return("both")
  } else if (a1c & !ogtt) {
    return("a1c")
  } else if (ogtt & !a1c) {
    return("ogtt")
  } else if (!ogtt & !a1c) {
    return("neither")
  }
})
```

@@

```{r}
p <- prev_year_age %>%
  pivot_longer(`[10,19)`:`[31, Inf)`, names_to = "Age Group") %>%
  rename(Year = ReviewYear)
p$name <- "Prevalence of CFRD"
i <- incidence %>%
  select(
    Year, `Total Incidence Age 10-18`, `Total Incidence Age 19-30`,
    `Total Incidence Age >= 31`
  ) %>%
  pivot_longer(`Total Incidence Age 10-18`:`Total Incidence Age >= 31`,
    names_to = "Age Group"
  )
i$name <- "Incidence of CFRD"
s <- screening %>%
  filter(name == "Screened OGTT") %>%
  select(ReviewYear, age_group, value) %>%
  rename(Year = ReviewYear, `Age Group` = age_group)
s$name <- "Screening Rates for CFRD"
s$value <- s$value * 100
t <- rbind(p, i, s)
t$`Age Group` <- factor(t$`Age Group`,
  levels = c(
    "[10,19)", "[19,31)", "[31, Inf)", "Total Incidence Age >= 31",
    "Total Incidence Age 10-18", "Total Incidence Age 19-30"
  ),
  labels = c(
    "10-18", "19-30", ">= 31", ">= 31",
    "10-18", "19-30"
  )
)
t <- t[t$Year >= 2004, ]
# Figure 1
fig1a <- ggplot(
  t[t$name == "Screening Rates for CFRD", ],
  aes(x = Year, y = value, linetype = `Age Group`)
) +
  geom_line() +
  theme_bw() +
  ylab("Percentage") +
  ggtitle("Screening Rates for CFRD")
fig1b <- ggplot(
  t[t$name == "Incidence of CFRD", ],
  aes(x = Year, y = value, linetype = `Age Group`)
) +
  geom_line() +
  theme_bw() +
  ylab("Per 1,000 Person-Years") +
  ggtitle("Incidence of CFRD")
fig1c <- ggplot(
  t[t$name == "Prevalence of CFRD", ],
  aes(x = Year, y = value, linetype = `Age Group`)
) +
  geom_line() +
  theme_bw() +
  ylab("Per 1,000 Person-Years") +
  ggtitle("Prevalence of CFRD")
fig1 <- fig1a + fig1b + fig1c + plot_layout(guides = "collect") +
  plot_annotation(title = "Figure 1: Trends in Screening, Incidence, and Prevalence of CFRD from 2003-2018, by age group ")
fig1
ggsave("./Dissemination/Figures/fig1.png",
  plot = fig1, width = 12, height = 3, units = "in", dpi = 900
)
# Graphical abstract is the same but with color
fig1a <- ggplot(
  t[t$name == "Screening Rates for CFRD", ],
  aes(x = Year, y = value, color = `Age Group`)
) +
  geom_line() +
  theme_bw() +
  ylab("Percentage") +
  ggtitle("Screening Rates for CFRD")
fig1b <- ggplot(
  t[t$name == "Incidence of CFRD", ],
  aes(x = Year, y = value, color = `Age Group`)
) +
  geom_line() +
  theme_bw() +
  ylab("Per 1,000 Person-Years") +
  ggtitle("Incidence of CFRD")
fig1c <- ggplot(
  t[t$name == "Prevalence of CFRD", ],
  aes(x = Year, y = value, color = `Age Group`)
) +
  geom_line() +
  theme_bw() +
  ylab("Per 1,000 Person-Years") +
  ggtitle("Prevalence of CFRD")
fig1 <- fig1a + fig1b + fig1c + plot_layout(guides = "collect")+
  plot_annotation(title = "US Cystic Fibrosis Foundation Patient Registry Trends in Screening, Incidence, and Prevalence of Cystic Fibrosis Related Diabetes from 2003-2018")
fig1
ggsave("./Dissemination/Figures/graphical_abstract.png",
  plot = fig1, width = 13, height = 3, units = "in", dpi = 900
)
```

@@

```{r}
# Origin for survival analyses
og <- ymd("2003-01-01")
# Need to use a dataframe not a tibble
annual <- as.data.frame(annual)
annual$tstart <- as.numeric(annual$tstart) - as.numeric(og)
annual$tstop <- as.numeric(annual$tstop) - as.numeric(og)
# Formula for base model
xvars <- c(
  "age_group", "Gender", "Race", "Hispanicrace", "bmi_cat", "MutClass",
  "A_IsOnEnzymes", "tube_feeding", "liver_disease", "ppFEV1_cat",
  "NumPulmExacerbation", "A_corticosteroids1"
)
y <- "Surv(time = tstart, time2 = tstop, event = as.numeric(CFRD) == 2)"
f <- as.formula(paste0(y, "~", paste0(xvars, collapse = "+")))
cox_res <- coxph(f, data = annual, cluster = eDWID)
# Forest plot and coefficients table
cox_res %>%
  tbl_regression(exponentiate = T)
ggcoef_table(cox_res,
  show_p_values = F, add_reference_rows = T, exponentiate = T,
  signif_stars = F, colour = F
)
# Diagnostics
test.ph <- data.frame(cox.zph(cox_res)$table)
test.ph <- test.ph[rownames(test.ph) != "GLOBAL", ]
# Find which variables technically violate PH assumption
time_varying <- rownames(test.ph)[test.ph$p < 0.05]
# Just split data by time period so we're only fitting 3 models
# "[2003,2009)", "[2009,2014)", "[2014,2019)"
annual$time_period <- cut(annual$ReviewYear,
  breaks = c(-Inf, 2009, 2014, Inf), right = F
)
era1 <- coxph(f,
  data = annual[as.numeric(annual$time_period) == 1, ],
  cluster = eDWID
)
era2 <- coxph(f,
  data = annual[as.numeric(annual$time_period) == 2, ],
  cluster = eDWID
)
era3 <- coxph(f,
  data = annual[as.numeric(annual$time_period) == 3, ],
  cluster = eDWID
)
# # KM plot
# fit <- survfit(Surv(time = tstart, time2 = tstop, event = as.numeric(CFRD) == 2) ~ age_group,
#   data = annual[as.numeric(annual$time_period) == 1, ]
# )
# # Visualize with survminer
# ggsurvplot(fit, data = annual[as.numeric(annual$time_period) == 1, ], risk.table = TRUE)
```

@@

```{r results='asis'}
f_vx770 <- update(f, . ~ . + A_Vx770)
vx770_cox_res <- coxph(f_vx770,
  data = annual[annual$ReviewYear >= 2011, ],
  cluster = eDWID
)
t1 <- vx770_cox_res %>%
  tbl_regression(exponentiate = T)
t1
ggcoef_table(vx770_cox_res,
  show_p_values = F, add_reference_rows = T, exponentiate = T,
  signif_stars = F, colour = F
)
```

@@

```{r}
f_vx809comb <- update(f, . ~ . + A_VX809comb)
vx809comb_cox_res <- coxph(f_vx809comb,
  data = annual[annual$ReviewYear >= 2015, ],
  cluster = eDWID
)
t2 <- vx809comb_cox_res %>%
  tbl_regression(exponentiate = T)
t2
ggcoef_table(vx809comb_cox_res,
  show_p_values = F, add_reference_rows = T, exponentiate = T,
  signif_stars = F, colour = F
)
```

>>>>

```{r}
#| include: false
library(Hmisc)
library(zoo)
library(tidyverse)
library(childsds)
library(gt)
library(gtsummary)
library(forcats)
library(labelled)
library(survival)
library(survminer)
library(knitr)
library(eulerr)
library(broom)
library(emmeans)
library(ggstats)
library(patchwork)
home_dir <- switch(Sys.info()["sysname"],
  "Darwin" = "/Users/tim/Library/CloudStorage/OneDrive-TheUniversityofColoradoDenver/Vigers/CF/Christine Chan/CF Registry",
  "Linux" = "/home/timvigers/OneDrive/Vigers/CF/Christine Chan/CF Registry"
)
github_dir <- switch(Sys.info()["sysname"],
  "Darwin" = "/Users/tim/Documents/GitHub",
  "Linux" = "/home/tim/Documents/GitHub"
)
knitr::opts_knit$set(root.dir = home_dir)
```

@@

  "A_pulmonarycomplications1", "A_mycocultureresults3", "A_cultureresults3"

@@

```{r}
# Per reviewer, out of X individuals with a diagnosis of CFRD, X number had an
# OGTT (and X had an A1c, but no OGTT) that year or the year prior.
# This is super slow but idc idc
first_cfrd=head(first_cfrd,100)
t <- apply(first_cfrd, 1, function(r) {
  id <- as.numeric(r["eDWID"])
  year <- as.numeric(r["first_cfrd"])
  d <- annual[annual$eDWID == id & annual$ReviewYear %in% c(year, year - 1), ]
  if (any(!is.na(d$A_hgba1c))) {
    a1c <- T
  } else {
    a1c <- F
  }
  if (any(!is.na(d["A_twohour_bloodglucose"]) |
    !is.na(d["A_ogttfast_bloodglucose"]))) {
    ogtt <- T
  } else {
    ogtt <- F
  }
  if (a1c & ogtt) {
    return("both")
  } else if (a1c & !ogtt) {
    return("a1c")
  } else if (ogtt & !a1c) {
    return("ogtt")
  } else if (!ogtt & !a1c) {
    return("neither")
  }
})
first_cfrd$a1c_ogtt <- t
cfrd_only <- left_join(first_cfrd, annual |> remove_labels(),
  by = join_by(eDWID, first_cfrd == ReviewYear)
)
cfrd_only$a1c_ogtt <- factor(cfrd_only$a1c_ogtt,
  levels = c("both", "ogtt", "a1c", "neither"),
  labels = c("OGTT", "OGTT", "HbA1c Only", "Neither")
)
cfrd_only %>%
  select(
    a1c_ogtt, Age_YrEnd, A_weight, A_height, A_bmivalue, A_bmipercentile,
    A_FEV1_pct_predicted, A_FVC_pct_predicted, Race, Hispanicrace, MutClass,
    A_hgba1c, A_ogttfast_bloodglucose, A_twohour_bloodglucose, A_IsOnEnzymes,
    liver_disease, A_cultureresults3
  ) %>%
  tbl_summary(by = a1c_ogtt, missing_text = "Missing") %>%
  add_overall() %>%
  add_p(test=list(all_continuous() ~ "oneway.test"))
```

@@

```{r}
p <- prev_year_age %>%
  pivot_longer(`[10,19)`:`[31, Inf)`, names_to = "Age Group") %>%
  rename(Year = ReviewYear)
p$name <- "Prevalence of CFRD"
i <- incidence %>%
  select(
    Year, `Total Incidence Age 10-18`, `Total Incidence Age 19-30`,
    `Total Incidence Age >= 31`
  ) %>%
  pivot_longer(`Total Incidence Age 10-18`:`Total Incidence Age >= 31`,
    names_to = "Age Group"
  )
i$name <- "Incidence of CFRD"
s <- screening %>%
  filter(name == "Screened OGTT") %>%
  select(ReviewYear, age_group, value) %>%
  rename(Year = ReviewYear, `Age Group` = age_group)
s$name <- "Screening Rates for CFRD"
s$value <- s$value * 100
t <- rbind(p, i, s)
t$`Age Group` <- factor(t$`Age Group`,
  levels = c(
    "[10,19)", "[19,31)", "[31, Inf)", "Total Incidence Age >= 31",
    "Total Incidence Age 10-18", "Total Incidence Age 19-30"
  ),
  labels = c(
    "10-18", "19-30", ">= 31", ">= 31",
    "10-18", "19-30"
  )
)
t <- t[t$Year >= 2004, ]
# Figure 1
fig1a <- ggplot(
  t[t$name == "Screening Rates for CFRD", ],
  aes(x = Year, y = value, linetype = `Age Group`)
) +
  geom_line() +
  theme_bw() +
  ylab("Percentage") +
  ggtitle("Screening Rates for CFRD")
fig1b <- ggplot(
  t[t$name == "Incidence of CFRD", ],
  aes(x = Year, y = value, linetype = `Age Group`)
) +
  geom_line() +
  theme_bw() +
  ylab("Per 1,000 Person-Years") +
  ggtitle("Incidence of CFRD")
fig1c <- ggplot(
  t[t$name == "Prevalence of CFRD", ],
  aes(x = Year, y = value, linetype = `Age Group`)
) +
  geom_line() +
  theme_bw() +
  ylab("Per 1,000 Person-Years") +
  ggtitle("Prevalence of CFRD")
fig1 <- fig1a + fig1b + fig1c + plot_layout(guides = "collect") +
  plot_annotation(title = "Figure 1: Trends in Screening, Incidence, and Prevalence of CFRD from 2003-2018, by age group ")
fig1
ggsave("./Dissemination/Figures/fig1.png",
  plot = fig1, width = 12, height = 3, units = "in", dpi = 900
)
# Graphical abstract is the same but with color
fig1a <- ggplot(
  t[t$name == "Screening Rates for CFRD", ],
  aes(x = Year, y = value, color = `Age Group`)
) +
  geom_line() +
  theme_bw() +
  ylab("Percentage") +
  ggtitle("Screening Rates for CFRD")
fig1b <- ggplot(
  t[t$name == "Incidence of CFRD", ],
  aes(x = Year, y = value, color = `Age Group`)
) +
  geom_line() +
  theme_bw() +
  ylab("Per 1,000 Person-Years") +
  ggtitle("Incidence of CFRD")
fig1c <- ggplot(
  t[t$name == "Prevalence of CFRD", ],
  aes(x = Year, y = value, color = `Age Group`)
) +
  geom_line() +
  theme_bw() +
  ylab("Per 1,000 Person-Years") +
  ggtitle("Prevalence of CFRD")
fig1 <- fig1a + fig1b + fig1c + plot_layout(guides = "collect") +
  plot_annotation(title = "US Cystic Fibrosis Foundation Patient Registry Trends in Screening, Incidence, and Prevalence of Cystic Fibrosis Related Diabetes from 2003-2018")
fig1
ggsave("./Dissemination/Figures/graphical_abstract.png",
  plot = fig1, width = 13, height = 3, units = "in", dpi = 900
)
```

@@

```{r}
# Origin for survival analyses
og <- ymd("2003-01-01")
# Need to use a dataframe not a tibble
annual <- as.data.frame(annual)
annual$tstart <- as.numeric(annual$tstart) - as.numeric(og)
annual$tstop <- as.numeric(annual$tstop) - as.numeric(og)
# Write cleaned data
save(annual, file = "./Data_Clean/survival_dataset.RData")
# Formula for base model
xvars <- c(
  "age_group", "Gender", "Race", "Hispanicrace", "bmi_cat", "MutClass",
  "A_IsOnEnzymes", "tube_feeding", "liver_disease", "ppFEV1_cat",
  "NumPulmExacerbation", "A_corticosteroids1"
)
y <- "Surv(time = tstart, time2 = tstop, event = as.numeric(CFRD) == 2)"
f <- as.formula(paste0(y, "~", paste0(xvars, collapse = "+")))
cox_res <- coxph(f, data = annual, cluster = eDWID)
# Forest plot and coefficients table
cox_res %>%
  tbl_regression(exponentiate = T)
ggcoef_table(cox_res,
  show_p_values = F, add_reference_rows = T, exponentiate = T,
  signif_stars = F, colour = F
)
# Diagnostics
cox.zph(cox_res)
# Just split data by time period so we're only fitting 3 models
# "[2003,2009)", "[2009,2014)", "[2014,2019)"
annual$time_period <- cut(annual$ReviewYear,
  breaks = c(-Inf, 2009, 2014, Inf), right = F
)
era1 <- coxph(f,
  data = annual[as.numeric(annual$time_period) == 1, ],
  cluster = eDWID
)
era2 <- coxph(f,
  data = annual[as.numeric(annual$time_period) == 2, ],
  cluster = eDWID
)
era3 <- coxph(f,
  data = annual[as.numeric(annual$time_period) == 3, ],
  cluster = eDWID
)
# # KM plot
# fit <- survfit(Surv(time = tstart, time2 = tstop, event = as.numeric(CFRD) == 2) ~ age_group,
#   data = annual[as.numeric(annual$time_period) == 1, ]
# )
# # Visualize with survminer
# ggsurvplot(fit, data = annual[as.numeric(annual$time_period) == 1, ], risk.table = TRUE)
```

@@

```{r results='asis'}
f_vx770 <- update(f, . ~ . + A_Vx770 - A_corticosteroids1 - NumPulmExacerbation - age_group)
vx770_cox_res <- coxph(f_vx770,
  data = annual[annual$ReviewYear >= 2011, ],
  cluster = eDWID
)
t1 <- vx770_cox_res %>%
  tbl_regression(exponentiate = T)
t1
ggcoef_table(vx770_cox_res,
  show_p_values = F, add_reference_rows = T, exponentiate = T,
  signif_stars = F, colour = F
)
```

@@

```{r}
f_vx809comb <- update(f, . ~ . + A_VX809comb - A_corticosteroids1 - NumPulmExacerbation - age_group - Hispanicrace)
vx809comb_cox_res <- coxph(f_vx809comb,
  data = annual[annual$ReviewYear >= 2015, ],
  cluster = eDWID
)
t2 <- vx809comb_cox_res %>%
  tbl_regression(exponentiate = T)
t2
ggcoef_table(vx809comb_cox_res,
  show_p_values = F, add_reference_rows = T, exponentiate = T,
  signif_stars = F, colour = F
)
```
